# Evaluation against drugs with TB HDT evidence in the literature
# Pubmed search on all FDA-approved drugs performed on April 2nd, 2026 by Ling Thang
# This notebook contains step-by-step evaluation of the top 20 drugs predicted using (i) our integrative approach, (2) the six connectivity methods across microarray/RNAseq individual signatures and aggregated signatures
# Last modified: 04/03/26
# Kewalin Samart

```{r}
library(readr)
library(here)
```

```{r}
labels <- read_tsv(here("evaluation/labels/launched_drug_with_literature.tsv"))
# HDT_label = 1 for Direct (HDT) matches.
# exception: "fostaminitib" is manually labeled as 1 due to a drug name typo
# in the literature (misspelled drug name in abstract prevented detection)
# https://doi.org/10.3390/antibiotics9090550
labels$HDT_label <- ifelse(
  labels$search_type == "Direct (HDT)" | labels$drug_name %in% c("fostamatinib"),
  1, 0
)
```

```{r}
# read in RNAseq indiv signatures predictions by method
CMAP_indivSig_TB_RNAseq <- readRDS(here("results/RNAseq/04_rank_aggregation/CMAP_indivSig_TB_RNAseq.rds"))
Cor_pearson_indivSig_TB_RNAseq <- readRDS(here("results/RNAseq/04_rank_aggregation/Cor_pearson_indivSig_TB_RNAseq.rds"))
Cor_spearman_indivSig_TB_RNAseq <- readRDS(here("results/RNAseq/04_rank_aggregation/Cor_spearman_indivSig_TB_RNAseq.rds"))
NCS_indivSig_TB_RNAseq <- readRDS(here("results/RNAseq/04_rank_aggregation/NCS_indivSig_TB_RNAseq.rds"))
Tau_indivSig_TB_RNAseq <- readRDS(here("results/RNAseq/04_rank_aggregation/Tau_indivSig_TB_RNAseq.rds"))
WCS_indivSig_TB_RNAseq <- readRDS(here("results/RNAseq/04_rank_aggregation/WCS_indivSig_TB_RNAseq.rds"))

# read in microarray indiv signatures predictions by method
CMAP_indivSig_TB_microarray <- readRDS(here("results/microarray/04_rank_aggregation/CMAP_indivSig_TB_microarray.rds"))
Cor_pearson_indivSig_TB_microarray <- readRDS(here("results/microarray/04_rank_aggregation/Cor_pearson_indivSig_TB_microarray.rds"))
Cor_spearman_indivSig_TB_microarray <- readRDS(here("results/microarray/04_rank_aggregation/Cor_spearman_indivSig_TB_microarray.rds"))
NCS_indivSig_TB_microarray <- readRDS(here("results/microarray/04_rank_aggregation/NCS_indivSig_TB_microarray.rds"))
Tau_indivSig_TB_microarray <- readRDS(here("results/microarray/04_rank_aggregation/Tau_indivSig_TB_microarray.rds"))
WCS_indivSig_TB_microarray <- readRDS(here("results/microarray/04_rank_aggregation/WCS_indivSig_TB_microarray.rds"))

# read in RNAseq aggregated signatures predictions by method
allscores_aggrSig_RNAseq <- readRDS(here("results/RNAseq/04_rank_aggregation/ranked_aggrSig_drugs_scores_neg_left_0.9_RNAseq_list.rds"))

# read in microarray aggregated signatures predictions by method
allscores_aggrSig_microarray <- readRDS(here("results/microarray/04_rank_aggregation/ranked_aggrSig_drugs_scores_neg_left_0.9_microarray_list.rds"))
```

```{r}
method_lists <- list(
  CMAP_microarray = CMAP_indivSig_TB_microarray$ranked_pert[1:20],
  CMAP_RNAseq     = CMAP_indivSig_TB_RNAseq$ranked_pert[1:20],
  CMAP_microarray_aggr = allscores_aggrSig_microarray$CMAP[1:20],
  CMAP_RNAseq_aggr = allscores_aggrSig_RNAseq$CMAP[1:20],

  Pearson_microarray = Cor_pearson_indivSig_TB_microarray$ranked_pert[1:20],
  Pearson_RNAseq     = Cor_pearson_indivSig_TB_RNAseq$ranked_pert[1:20],
  Pearson_microarray_aggr = allscores_aggrSig_microarray$Cor_pearson[1:20],
  Pearson_RNAseq_aggr = allscores_aggrSig_RNAseq$Cor_pearson[1:20],

  Spearman_microarray = Cor_spearman_indivSig_TB_microarray$ranked_pert[1:20],
  Spearman_RNAseq     = Cor_spearman_indivSig_TB_RNAseq$ranked_pert[1:20],
  Spearman_microarray_aggr = allscores_aggrSig_microarray$Cor_spearman[1:20],
  Spearman_RNAseq_aggr = allscores_aggrSig_RNAseq$Cor_spearman[1:20],

  NCS_microarray = NCS_indivSig_TB_microarray$ranked_pert[1:20],
  NCS_RNAseq     = NCS_indivSig_TB_RNAseq$ranked_pert[1:20],
  NCS_microarray_aggr = allscores_aggrSig_microarray$NCS[1:20],
  NCS_RNAseq_aggr = allscores_aggrSig_RNAseq$NCS[1:20],

  Tau_microarray = Tau_indivSig_TB_microarray$ranked_pert[1:20],
  Tau_RNAseq     = Tau_indivSig_TB_RNAseq$ranked_pert[1:20],
  Tau_microarray_aggr = allscores_aggrSig_microarray$Tau[1:20],
  Tau_RNAseq_aggr = allscores_aggrSig_RNAseq$Tau[1:20],

  WCS_microarray = WCS_indivSig_TB_microarray$ranked_pert[1:20],
  WCS_RNAseq     = WCS_indivSig_TB_RNAseq$ranked_pert[1:20],
  WCS_microarray_aggr = allscores_aggrSig_microarray$WCS[1:20],
  WCS_RNAseq_aggr = allscores_aggrSig_RNAseq$WCS[1:20]
)

integrative_pred = read_tsv(here("results/final_high_confidence_drug_predictions.tsv"))
integrative_pred_list <- integrative_pred$drug_name[1:20]

method_lists$Integrative <- integrative_pred_list
```

```{r}
# collect all predicted drugs (top 20 single-score lists + integrative)
all_predicted_drugs <- unique(unlist(method_lists))

# restrict labels to evaluated drugs only
labels_eval <- all_predicted_drugs[all_predicted_drugs %in% labels$drug_name]
all_labeled <- all(all_predicted_drugs %in% labels$drug_name)

print(paste0(
  "Do all predicted drugs have labels? ",
  all_labeled
))

missing_drugs <- setdiff(all_predicted_drugs, labels$drug_name)

print(paste0("Missing label for:", missing_drugs))

# count positives in evaluated set
num_pos <- sum(labels[labels$drug_name %in% labels_eval,]$HDT_label, na.rm = TRUE)
print(paste0("Size of evaluated set (top 20 predictions from each list): ", length(labels[labels$drug_name %in% labels_eval,]$HDT_label)))
print(paste0("Number of positives in evaluated set (among top 20 predictions from each list): ", num_pos))
print(paste0("Number of all positives: ", sum(labels$HDT_label)))

```

```{r}
compute_metrics <- function(drug_list, positives, K) {
  drug_list <- tolower(drug_list)
  
  topk <- drug_list[1:K]
  
  hits <- sum(topk %in% positives)
  
  precision <- hits / K
  recall <- hits / length(positives)
  
  c(Precision = precision, Recall = recall)
}
```

```{r}
labels$drug_name <- tolower(labels$drug_name)
positives <- labels$drug_name[labels$HDT_label == 1]

```

```{r}
library(dplyr)

rank_df <- lapply(names(method_lists), function(method) {
  
  drugs <- tolower(method_lists[[method]])
  
  data.frame(
    Method = method,
    drug_name = drugs,
    Rank = seq_along(drugs),
    signature_type = ifelse(grepl("aggr", method, ignore.case = TRUE),
                            "Aggregated signatures",
                            "Individual signatures")
  )
}) %>%
  bind_rows()

# keep only known positives
rank_df <- rank_df %>%
  filter(drug_name %in% positives)
rank_df$Score <- gsub("_.*", "", rank_df$Method)

```

```{r}
library(ggplot2)
library(dplyr)

# order Score by median rank (lower = better)
rank_df <- rank_df %>%
  mutate(Score = gsub("_.*", "", Method)) %>%
  group_by(Score) %>%
  mutate(median_rank = median(Rank, na.rm = TRUE)) %>%
  ungroup()

rank_df$Score <- reorder(rank_df$Score, rank_df$median_rank)

ggplot(rank_df, aes(x = Score, y = Rank)) +
  geom_boxplot(alpha = 0.7) +
  scale_y_reverse() +  # rank 1 at top
  coord_flip() +       # flip axes
  theme_minimal() +
  labs(
    title = "Distribution of ranks for known drugs in the top 20 predictions across methods",
    x = "Method",
    y = "Rank (Top = 1)"
  ) +
  theme(legend.position = "none")
```

```{r}
library(dplyr)
library(ggplot2)

rank_df_expanded <- rank_df %>%
  mutate(signature_type = ifelse(Method == "Integrative", signature_type, signature_type)) %>%
  # add "All" panel for Integrative
  bind_rows(
    rank_df %>%
      filter(Method == "Integrative") %>%
      mutate(signature_type = "All (Integrative)")
  )

# reorder signature_type by median rank (ascending)
rank_df_expanded <- rank_df_expanded %>%
  mutate(signature_type = factor(signature_type, 
                                 levels = names(sort(tapply(Rank, signature_type, median)))))

ggplot(rank_df_expanded, aes(x = signature_type, y = Rank, fill = signature_type)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +    
  scale_y_reverse() +   # rank 1 at top
  coord_flip() +        # horizontal boxes
  theme_minimal() +
  labs(
    title = "Distribution of ranks for known drugs in \n the top 20 predictions by signature type",
    x = "Signature type",
    y = "Rank (Top = 1)"
  ) +
  theme(legend.position = "none")
```

```{r}
library(dplyr)

rank_df_plot <- rank_df %>%
  bind_rows(
    rank_df %>%
      filter(Score == "Integrative") %>%
      mutate(DataType = "RNAseq"),
    rank_df %>%
      filter(Score == "Integrative") %>%
      mutate(DataType = "Microarray")
  )

library(dplyr)
library(tidyr)

# extract Integrative rows
integrative_rows <- rank_df %>%
  filter(Score == "Integrative") %>%
  select(-DataType) %>%   # remove the existing DataType column
  crossing(DataType = c("RNAseq", "Microarray"))

# combine with the rest (excluding original Integrative)
rank_df_plot <- rank_df %>%
  filter(Score != "Integrative") %>%
  bind_rows(integrative_rows)

ggplot(rank_df_plot, aes(
  x = Score,
  y = Rank,
  fill = signature_type,
  alpha = Score == "Integrative"  # TRUE/FALSE
)) +
  geom_boxplot(color = "black") +
  scale_y_reverse() +
  coord_flip() +
  facet_wrap(~DataType) +
  theme_minimal() +
  scale_fill_manual(
    values = c(
      "Aggregated signatures" = "#0072B2",
      "Individual signatures" = "#D55E00"
    )
  ) +
  scale_alpha_manual(
    values = c(`TRUE` = 0, `FALSE` = 0.8),  # Integrative vs others
    guide = "none"  
  ) +
  labs(
    title = "Distribution of ranks for known drugs in the top 20 predictions across methods",
    x = "Method",
    y = "Rank (Top = 1)",
    fill = "Signature Type"
  ) +
  theme(
    legend.position = "right",
    strip.text = element_text(face = "bold")
  )

```
```{r}
# set the order of methods
rank_df_plot$Score <- factor(rank_df_plot$Score,
                             levels = rev(c("CMAP", "WCS", "NCS", "Tau", "Pearson", "Spearman", "Integrative")))

# calculate median rank for Integrative
median_integrative <- rank_df_plot %>%
  filter(Score == "Integrative") %>%
  summarise(med = median(Rank)) %>%
  pull(med)

# Plot
ggplot(rank_df_plot, aes(
  x = Score,
  y = Rank,
  fill = signature_type,
  alpha = Score == "Integrative"  # TRUE/FALSE
)) +
  geom_boxplot(color = "black") +
  # ddd horizontal line at median of Integrative (vertical after flip)
  geom_hline(yintercept = median_integrative, linetype = "dashed", color = "red", size = 0.5) +
  scale_y_reverse() +
  coord_flip() +
  facet_wrap(~DataType) +
  theme_minimal() +
  scale_fill_manual(
    values = c(
      "Aggregated signatures" = "#0072B2",
      "Individual signatures" = "#D55E00"
    )
  ) +
  scale_alpha_manual(
    values = c(`TRUE` = 0, `FALSE` = 0.8),  # Integrative vs others
    guide = "none"  # hide alpha legend
  ) +
  labs(
    title = "Distribution of ranks for known drugs in the top 20 predictions across methods",
    x = "Method",
    y = "Rank (Top = 1)",
    fill = "Signature Type"
  ) +
  theme(
    legend.position = "right",
    strip.text = element_text(face = "bold")
  )

```



```{r}
rank_df <- rank_df %>%
  mutate(
    Score = gsub("_.*", "", Method),
    DataType = ifelse(grepl("RNAseq", Method), "RNAseq",
                      ifelse(grepl("microarray", Method), "Microarray", "Integrative")),
    Method_full = ifelse(DataType == "Integrative",
                         "Integrative",
                         paste(Score, DataType, sep = "_"))
  )
rank_df <- rank_df %>%
  group_by(Method_full) %>%
  mutate(median_rank = median(Rank, na.rm = TRUE)) %>%
  ungroup()

rank_df$Method_full <- reorder(rank_df$Method_full, rank_df$median_rank)
```

```{r}
library(ggplot2)
library(dplyr)

rank_df <- rank_df %>%
  mutate(fill_color = ifelse(Method == "Integrative", "black",
                             ifelse(signature_type == "Aggregated signatures", "#1f77b4",
                                    "#ff7f0e")))

ggplot(rank_df, aes(x = Method_full, y = Rank, fill = fill_color)) +
  geom_boxplot(
    alpha = 0.7,
    position = position_dodge(width = 0.8),
    outlier.shape = NA
  ) +
  scale_fill_identity() +  
  scale_y_reverse() +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Distribution of ranks for known drugs \n by method-signature-type",
    x = "Method",
    y = "Rank (Top = 1)"
  ) +
  theme(legend.position = "right")
```

```{r}
Ks <- c(5, 10, 15, 20)

results <- lapply(Ks, function(k) {
  res <- lapply(method_lists, function(dl) {
    compute_metrics(dl, positives, K = k)
  })
  
  df <- as.data.frame(do.call(rbind, res))
  df$Method <- names(method_lists)
  df$K <- k
  df
})

results_df <- do.call(rbind, results)
```

```{r}
library(dplyr)
library(ggplot2)
library(RColorBrewer)

results_df$DataType <- ifelse(grepl("RNAseq", results_df$Method), "RNAseq", "Microarray")

# add signature_type column
results_df <- results_df %>%
  mutate(signature_type = case_when(
    grepl("ggr", Method) ~ "Aggregated signature",
    grepl("Integrative", Method) ~ "Integrative",
    TRUE ~ "Individual signature"
  ))

results_df$Score <- gsub("_.*", "", results_df$Method)

results_df$IsIntegrative <- grepl("Integrative", results_df$Method)

score_colors <- c(
  "Integrative" = "black",
  setNames(brewer.pal(length(unique(results_df$Score)), "Dark2"),
           unique(results_df$Score))
)

# Plot
ggplot(results_df, aes(x = K, y = Precision, group = Method)) +
  geom_line(aes(color = Score, linetype = signature_type), linewidth = 1, alpha = 0.7) +
  geom_point(aes(
    shape = ifelse(IsIntegrative, "Integrative", DataType),
    color = ifelse(IsIntegrative, "Integrative", Score)
  ), size = 3) +
  scale_color_manual(values = score_colors) +
  scale_shape_manual(values = c(
    "Integrative" = 15,   # square
    "RNAseq" = 16,        # circle
    "Microarray" = 17     # triangle
  )) +
  scale_linetype_manual(values = c(
    "Individual signature" = "dashed",
    "Aggregated signature" = "solid",
    "Integrative" = "solid"
  )) +
  theme_minimal() +
  labs(
    title = "Precision@K across method-technology combinations",
    x = "K",
    y = "Precision",
    color = "Score",
    shape = "Data type",
    linetype = "Signature type"
  )
```

```{r}
library(dplyr)

results_df_plot <- results_df %>%
  bind_rows(
    results_df %>%
      filter(grepl("Integrative", Method)) %>%
      mutate(DataType = "RNAseq")
  )

ggplot(results_df_plot, aes(x = K, y = Precision, group = Method)) +
  
  geom_line(
    data = subset(results_df_plot, !grepl("Integrative", Method)),
    aes(color = Score),
    linewidth = 1,
    alpha = 0.5
  ) +
  geom_point(
    data = subset(results_df_plot, !grepl("Integrative", Method)),
    aes(color = Score),
    size = 3,
    alpha = 0.5
  ) +
  
  geom_line(
    data = subset(results_df_plot, grepl("Integrative", Method)),
    aes(color = Score),
    linewidth = 1.5,
    alpha = 1
  ) +
  geom_point(
    data = subset(results_df_plot, grepl("Integrative", Method)),
    aes(color = Score),
    size = 3.5,
    alpha = 1
  ) +
  
  facet_wrap(~ DataType) +
  scale_color_manual(values = score_colors) +
  theme_minimal() +
  labs(
    title = "Precision@K across methods by technology",
    x = "K",
    y = "Precision",
    color = "Score"
  )

```

```{r}
library(dplyr)
library(ggplot2)
library(RColorBrewer)

# add signature_type
results_df <- results_df %>%
  mutate(
    DataType = ifelse(grepl("RNAseq", Method), "RNAseq", "Microarray"),
    signature_type = case_when(
      grepl("ggr", Method) ~ "Aggregated signature",
      grepl("Integrative", Method) ~ "Integrative",
      TRUE ~ "Individual signature"
    ),
    Score = gsub("_.*", "", Method),
    IsIntegrative = grepl("Integrative", Method)
  )

results_df_plot <- results_df %>%
  bind_rows(
    results_df %>%
      filter(IsIntegrative) %>%
      mutate(DataType = "RNAseq")
  )

score_colors <- c(
  "Integrative" = "black",
  setNames(brewer.pal(length(unique(results_df$Score)), "Dark2"),
           unique(results_df$Score))
)


ggplot(results_df_plot, aes(x = K, y = Precision, group = Method)) +
  
  geom_line(
    data = subset(results_df_plot, !IsIntegrative),
    aes(color = Score, linetype = signature_type),
    linewidth = 1,
    alpha = 0.5
  ) +
  geom_point(
    data = subset(results_df_plot, !IsIntegrative),
    aes(color = Score),
    size = 3,
    alpha = 0.5
  ) +
  
  geom_line(
    data = subset(results_df_plot, IsIntegrative),
    aes(color = Score, linetype = signature_type),
    linewidth = 1.5,
    alpha = 1
  ) +
  geom_point(
    data = subset(results_df_plot, IsIntegrative),
    aes(color = Score),
    size = 3.5,
    alpha = 1
  ) +
  
  facet_wrap(~ DataType) +
  
  scale_color_manual(values = score_colors) +
  scale_linetype_manual(values = c(
    "Individual signature" = "dashed",
    "Aggregated signature" = "solid",
    "Integrative" = "solid"
  )) +
  
  theme_minimal() +
  labs(
    title = "Precision@K across methods by technology",
    x = "K",
    y = "Precision",
    color = "Score",
    linetype = "Signature Type"
  )
```

```{r}
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(tidyr)

# add signature_type and IsIntegrative
results_df <- results_df %>%
  mutate(
    DataType = ifelse(grepl("RNAseq", Method), "RNAseq", "Microarray"),
    signature_type = case_when(
      grepl("ggr", Method) ~ "Aggregated signature",
      grepl("Integrative", Method) ~ "Aggregated signature", # Integrative treated as solid
      TRUE ~ "Individual signature"
    ),
    Score = gsub("_.*", "", Method),
    IsIntegrative = grepl("Integrative", Method)
  )

# duplicate Integrative rows into all facets without adding a new facet row
integrative_rows <- results_df %>%
  filter(IsIntegrative) %>%
  select(-DataType, -signature_type) %>%  # remove to avoid duplication errors
  crossing(
    DataType = c("RNAseq", "Microarray"),
    signature_type = c("Aggregated signature", "Individual signature") # Integrative appears in both facets
  )

results_df_plot <- bind_rows(results_df, integrative_rows)

# Define colors
score_colors <- c(
  "Integrative" = "black",
  setNames(brewer.pal(length(unique(results_df$Score)), "Dark2"),
           unique(results_df$Score))
)

# Plot
ggplot(results_df_plot, aes(x = K, y = Precision, group = Method)) +
  
  geom_line(
    data = subset(results_df_plot, !IsIntegrative),
    aes(color = Score, linetype = signature_type),
    linewidth = 1,
    alpha = 0.5
  ) +
  geom_point(
    data = subset(results_df_plot, !IsIntegrative),
    aes(color = Score),
    size = 3,
    alpha = 0.5
  ) +
  
  geom_line(
    data = subset(results_df_plot, IsIntegrative),
    aes(color = Score),
    linewidth = 1.5,
    linetype = "solid",
    alpha = 1
  ) +
  geom_point(
    data = subset(results_df_plot, IsIntegrative),
    aes(color = Score),
    size = 3.5,
    alpha = 1
  ) +
  
  facet_grid(signature_type ~ DataType) +
  
  scale_color_manual(values = score_colors) +
  scale_linetype_manual(values = c(
    "Individual signature" = "dashed",
    "Aggregated signature" = "solid"
  )) +
  
  theme_minimal() +
  labs(
    title = "Precision@K across methods by technology and signature type",
    x = "K",
    y = "Precision",
    color = "Score",
    linetype = "Signature Type"
  ) +
  theme(
    strip.text = element_text(face = "bold")
  )
```

```{r}
library(dplyr)
library(ggplot2)
library(RColorBrewer)

# ensure signature_type exists
results_df <- results_df %>%
  mutate(
    signature_type = case_when(
      grepl("ggr", Method) ~ "Aggregated signature",
      grepl("Integrative", Method) ~ "Aggregated signature", # Integrative treated as Aggregated
      TRUE ~ "Individual signature"
    ),
    Score = gsub("_.*", "", Method)
  )

# compute average Precision@K by Score, signature_type, and K
avg_df <- results_df %>%
  group_by(Score, signature_type, K) %>%
  summarise(AveragePrecision = mean(Precision, na.rm = TRUE), .groups = "drop")

score_colors <- c(
  "Integrative" = "black",
  setNames(RColorBrewer::brewer.pal(length(unique(results_df$Score)), "Dark2"),
           unique(results_df$Score))
)

# Plot
ggplot(avg_df, aes(x = K, y = AveragePrecision, color = Score)) +
  geom_line(
    aes(linetype = signature_type, alpha = ifelse(Score == "Integrative", 1, 0.5)),
    linewidth = 1.2
  ) +
  geom_point(
    aes(alpha = ifelse(Score == "Integrative", 1, 0.5)),
    size = 3
  ) +
  scale_color_manual(values = score_colors) +
  scale_linetype_manual(values = c(
    "Individual signature" = "dashed",
    "Aggregated signature" = "solid"
  )) +
  scale_alpha_identity() +   # use the alpha values as is
  theme_minimal() +
  labs(
    title = "Average precision@K across microarray and RNAseq",
    x = "K",
    y = "Average precision",
    color = "Score",
    linetype = "Signature type"
  )
```


```{r}
# Plot recall
ggplot(results_df, aes(x = K, y = Recall, group = Method)) +
  geom_line(aes(color = Score, linetype = signature_type), linewidth = 1, alpha = 0.7) +
  geom_point(aes(
    shape = ifelse(IsIntegrative, "Integrative", DataType),
    color = ifelse(IsIntegrative, "Integrative", Score)
  ), size = 3) +
  scale_color_manual(values = score_colors) +
  scale_shape_manual(values = c(
    "Integrative" = 15,   # square
    "RNAseq" = 16,        # circle
    "Microarray" = 17     # triangle
  )) +
  scale_linetype_manual(values = c(
    "Individual signature" = "dashed",
    "Aggregated signature" = "solid",
    "Integrative" = "solid"
  )) +
  theme_minimal() +
  labs(
    title = "Recall@K across method-technology combinations",
    x = "K",
    y = "Recall",
    color = "Score",
    shape = "Data type",
    linetype = "Signature type"
  )
```

```{r}
# Plot recall
ggplot(results_df_plot, aes(x = K, y = Recall, group = Method)) +
  
  # non-Integrative (faded, linetype by signature_type)
  geom_line(
    data = subset(results_df_plot, !IsIntegrative),
    aes(color = Score, linetype = signature_type),
    linewidth = 1,
    alpha = 0.5
  ) +
  geom_point(
    data = subset(results_df_plot, !IsIntegrative),
    aes(color = Score),
    size = 3,
    alpha = 0.5
  ) +
  
  geom_line(
    data = subset(results_df_plot, IsIntegrative),
    aes(color = Score, linetype = signature_type),
    linewidth = 1.5,
    alpha = 1
  ) +
  geom_point(
    data = subset(results_df_plot, IsIntegrative),
    aes(color = Score),
    size = 3.5,
    alpha = 1
  ) +
  
  facet_wrap(~ DataType) +
  
  scale_color_manual(values = score_colors) +
  scale_linetype_manual(values = c(
    "Individual signature" = "dashed",
    "Aggregated signature" = "solid",
    "Integrative" = "solid"
  )) +
  
  theme_minimal() +
  labs(
    title = "Recall@K across methods by technology",
    x = "K",
    y = "Recall",
    color = "Score",
    linetype = "Signature Type"
  )

```

```{r}
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(tidyr)

# add signature_type and IsIntegrative
results_df <- results_df %>%
  mutate(
    DataType = ifelse(grepl("RNAseq", Method), "RNAseq", "Microarray"),
    signature_type = case_when(
      grepl("ggr", Method) ~ "Aggregated signature",
      grepl("Integrative", Method) ~ "Aggregated signature", # Integrative treated as solid
      TRUE ~ "Individual signature"
    ),
    Score = gsub("_.*", "", Method),
    IsIntegrative = grepl("Integrative", Method)
  )

# duplicate Integrative rows into all facets without adding a new facet row
integrative_rows <- results_df %>%
  filter(IsIntegrative) %>%
  select(-DataType, -signature_type) %>%  # remove to avoid duplication errors
  crossing(
    DataType = c("RNAseq", "Microarray"),
    signature_type = c("Aggregated signature", "Individual signature") # Integrative appears in both facets
  )

results_df_plot <- bind_rows(results_df, integrative_rows)

score_colors <- c(
  "Integrative" = "black",
  setNames(brewer.pal(length(unique(results_df$Score)), "Dark2"),
           unique(results_df$Score))
)

ggplot(results_df_plot, aes(x = K, y = Recall, group = Method)) +
  
  geom_line(
    data = subset(results_df_plot, !IsIntegrative),
    aes(color = Score, linetype = signature_type),
    linewidth = 1,
    alpha = 0.5
  ) +
  geom_point(
    data = subset(results_df_plot, !IsIntegrative),
    aes(color = Score),
    size = 3,
    alpha = 0.5
  ) +
  
  geom_line(
    data = subset(results_df_plot, IsIntegrative),
    aes(color = Score),
    linewidth = 1.5,
    linetype = "solid",
    alpha = 1
  ) +
  geom_point(
    data = subset(results_df_plot, IsIntegrative),
    aes(color = Score),
    size = 3.5,
    alpha = 1
  ) +
  
  facet_grid(signature_type ~ DataType) +
  
  scale_color_manual(values = score_colors) +
  scale_linetype_manual(values = c(
    "Individual signature" = "dashed",
    "Aggregated signature" = "solid"
  )) +
  
  theme_minimal() +
  labs(
    title = "Recall@K across methods by technology and signature type",
    x = "K",
    y = "Recall",
    color = "Score",
    linetype = "Signature type"
  ) +
  theme(
    strip.text = element_text(face = "bold")
  )

```

```{r}
library(dplyr)
library(ggplot2)
library(RColorBrewer)

# ensure signature_type exists
results_df <- results_df %>%
  mutate(
    signature_type = case_when(
      grepl("ggr", Method) ~ "Aggregated signature",
      grepl("Integrative", Method) ~ "Aggregated signature", # Integrative treated as Aggregated
      TRUE ~ "Individual signature"
    ),
    Score = gsub("_.*", "", Method)
  )

# compute average Precision@K by Score, signature_type, and K
avg_df_recall <- results_df %>%
  group_by(Score, signature_type, K) %>%
  summarise(AverageRecall = mean(Recall, na.rm = TRUE), .groups = "drop")

score_colors <- c(
  "Integrative" = "black",
  setNames(RColorBrewer::brewer.pal(length(unique(results_df$Score)), "Dark2"),
           unique(results_df$Score))
)

# Plot
ggplot(avg_df_recall, aes(x = K, y = AverageRecall, color = Score)) +
  geom_line(
    aes(linetype = signature_type, alpha = ifelse(Score == "Integrative", 1, 0.5)),
    linewidth = 1.2
  ) +
  geom_point(
    aes(alpha = ifelse(Score == "Integrative", 1, 0.5)),
    size = 3
  ) +
  scale_color_manual(values = score_colors) +
  scale_linetype_manual(values = c(
    "Individual signature" = "dashed",
    "Aggregated signature" = "solid"
  )) +
  scale_alpha_identity() +   
  theme_minimal() +
  labs(
    title = "Average recall@K across microarray and RNAseq",
    x = "K",
    y = "Average recall",
    color = "Score",
    linetype = "Signature type"
  )

```