# Evaluation against drugs with TB HDT evidence in the literature
# Pubmed search on all FDA-approved drugs performed on April 2nd, 2026 by Ling Thang
# This notebook contains step-by-step evaluation of the top 20 drugs predicted using (i) our integrative approach, (2) the six connectivity methods across microarray/RNAseq individual signatures and aggregated signatures
# Last modified: 04/09/26
# Kewalin Samart

```{r}
library(readr)
library(here)
```

```{r}
labels <- read_tsv(here("evaluation/labels/launched_drug_with_literature.tsv"))
# HDT_label = 1 for Direct (HDT) matches.
# exception: "fostaminitib" is manually labeled as 1 due to a drug name typo
# in the literature (misspelled drug name in abstract prevented detection)
# https://doi.org/10.3390/antibiotics9090550
labels$HDT_label <- ifelse(
  labels$search_type == "Direct (HDT)" | labels$drug_name %in% c("fostamatinib"),
  1, 0
)
```

```{r}
# read in RNAseq indiv signatures predictions by method
CMAP_indivSig_TB_RNAseq <- readRDS(here("results/RNAseq/04_rank_aggregation/CMAP_indivSig_TB_RNAseq.rds"))
Cor_pearson_indivSig_TB_RNAseq <- readRDS(here("results/RNAseq/04_rank_aggregation/Cor_pearson_indivSig_TB_RNAseq.rds"))
Cor_spearman_indivSig_TB_RNAseq <- readRDS(here("results/RNAseq/04_rank_aggregation/Cor_spearman_indivSig_TB_RNAseq.rds"))
NCS_indivSig_TB_RNAseq <- readRDS(here("results/RNAseq/04_rank_aggregation/NCS_indivSig_TB_RNAseq.rds"))
Tau_indivSig_TB_RNAseq <- readRDS(here("results/RNAseq/04_rank_aggregation/Tau_indivSig_TB_RNAseq.rds"))
WCS_indivSig_TB_RNAseq <- readRDS(here("results/RNAseq/04_rank_aggregation/WCS_indivSig_TB_RNAseq.rds"))

# read in microarray indiv signatures predictions by method
CMAP_indivSig_TB_microarray <- readRDS(here("results/microarray/04_rank_aggregation/CMAP_indivSig_TB_microarray.rds"))
Cor_pearson_indivSig_TB_microarray <- readRDS(here("results/microarray/04_rank_aggregation/Cor_pearson_indivSig_TB_microarray.rds"))
Cor_spearman_indivSig_TB_microarray <- readRDS(here("results/microarray/04_rank_aggregation/Cor_spearman_indivSig_TB_microarray.rds"))
NCS_indivSig_TB_microarray <- readRDS(here("results/microarray/04_rank_aggregation/NCS_indivSig_TB_microarray.rds"))
Tau_indivSig_TB_microarray <- readRDS(here("results/microarray/04_rank_aggregation/Tau_indivSig_TB_microarray.rds"))
WCS_indivSig_TB_microarray <- readRDS(here("results/microarray/04_rank_aggregation/WCS_indivSig_TB_microarray.rds"))

# read in RNAseq aggregated signatures predictions by method
allscores_aggrSig_RNAseq <- readRDS(here("results/RNAseq/04_rank_aggregation/ranked_aggrSig_drugs_scores_neg_left_0.9_RNAseq_list.rds"))

# read in microarray aggregated signatures predictions by method
allscores_aggrSig_microarray <- readRDS(here("results/microarray/04_rank_aggregation/ranked_aggrSig_drugs_scores_neg_left_0.9_microarray_list.rds"))
```

```{r}
method_lists <- list(
  CMAP_microarray = CMAP_indivSig_TB_microarray$ranked_pert[1:20],
  CMAP_RNAseq     = CMAP_indivSig_TB_RNAseq$ranked_pert[1:20],
  CMAP_microarray_aggr = allscores_aggrSig_microarray$CMAP[1:20],
  CMAP_RNAseq_aggr = allscores_aggrSig_RNAseq$CMAP[1:20],

  Pearson_microarray = Cor_pearson_indivSig_TB_microarray$ranked_pert[1:20],
  Pearson_RNAseq     = Cor_pearson_indivSig_TB_RNAseq$ranked_pert[1:20],
  Pearson_microarray_aggr = allscores_aggrSig_microarray$Cor_pearson[1:20],
  Pearson_RNAseq_aggr = allscores_aggrSig_RNAseq$Cor_pearson[1:20],

  Spearman_microarray = Cor_spearman_indivSig_TB_microarray$ranked_pert[1:20],
  Spearman_RNAseq     = Cor_spearman_indivSig_TB_RNAseq$ranked_pert[1:20],
  Spearman_microarray_aggr = allscores_aggrSig_microarray$Cor_spearman[1:20],
  Spearman_RNAseq_aggr = allscores_aggrSig_RNAseq$Cor_spearman[1:20],

  NCS_microarray = NCS_indivSig_TB_microarray$ranked_pert[1:20],
  NCS_RNAseq     = NCS_indivSig_TB_RNAseq$ranked_pert[1:20],
  NCS_microarray_aggr = allscores_aggrSig_microarray$NCS[1:20],
  NCS_RNAseq_aggr = allscores_aggrSig_RNAseq$NCS[1:20],

  Tau_microarray = Tau_indivSig_TB_microarray$ranked_pert[1:20],
  Tau_RNAseq     = Tau_indivSig_TB_RNAseq$ranked_pert[1:20],
  Tau_microarray_aggr = allscores_aggrSig_microarray$Tau[1:20],
  Tau_RNAseq_aggr = allscores_aggrSig_RNAseq$Tau[1:20],

  WCS_microarray = WCS_indivSig_TB_microarray$ranked_pert[1:20],
  WCS_RNAseq     = WCS_indivSig_TB_RNAseq$ranked_pert[1:20],
  WCS_microarray_aggr = allscores_aggrSig_microarray$WCS[1:20],
  WCS_RNAseq_aggr = allscores_aggrSig_RNAseq$WCS[1:20]
)

integrative_pred = read_tsv(here("results/final_high_confidence_drug_predictions.tsv"))
integrative_pred_list <- integrative_pred$drug_name[1:20]

method_lists$Integrative <- integrative_pred_list
```

```{r}
# predicted Drugs vs Labels
all_predicted_drugs <- unique(unlist(method_lists))
labels_eval <- all_predicted_drugs[all_predicted_drugs %in% labels$drug_name]
all_labeled <- all(all_predicted_drugs %in% labels$drug_name)

cat("Total unique predicted drugs:", length(all_predicted_drugs), "\n")
cat("All predicted drugs have labels:", all_labeled, "\n")

missing_drugs <- setdiff(all_predicted_drugs, labels$drug_name)
cat("Number of drugs missing labels:", length(missing_drugs), "\n")
if (length(missing_drugs) > 0) {
  cat("Missing drugs:\n")
  print(missing_drugs)
}

evaluated_labels <- labels[labels$drug_name %in% labels_eval, ]
num_pos <- sum(evaluated_labels$HDT_label, na.rm = TRUE)
total_eval <- nrow(evaluated_labels)
total_pos_all <- sum(labels$HDT_label, na.rm = TRUE)

cat("Size of evaluated set:", total_eval, "\n")
cat("Positives in evaluated set:", num_pos, "\n")
cat("Total positives in full dataset:", total_pos_all, "\n")

# drug metadata for FDA-approved labels
drug_rep <- read_csv(
  here("data/metadata/repurposing_drugs_20200324.csv"),
  skip = 9
)

drug_rep_filtered <- drug_rep %>%
  select(pert_iname, clinical_phase) %>%
  filter(clinical_phase == "Launched") %>%
  mutate(pert_iname = tolower(pert_iname))

cat("Launched drugs (rows):", nrow(drug_rep_filtered), "\n")
cat("Unique launched drugs:", length(unique(drug_rep_filtered$pert_iname)), "\n")

# LINCS filtering for specific condition: 24 h and 10 µM
lincs <- read_delim(
  here("data/metadata/GSE92742_Broad_LINCS_sig_info.txt"),
  delim = "\t"
)

lincs_filtered <- lincs %>%
  filter(
    pert_itime == "24 h",
    pert_idose == "10 µM"
  ) %>%
  mutate(pert_iname = tolower(pert_iname))

cat("Filtered LINCS profiles:", nrow(lincs_filtered), "\n")
cat("Unique LINCS drugs:", length(unique(lincs_filtered$pert_iname)), "\n")

# overlap between LINCS and FDA-approved drugs
overlap <- intersect(
  unique(lincs_filtered$pert_iname),
  unique(drug_rep_filtered$pert_iname)
)

cat("Number of overlapping drugs:", length(overlap), "\n")

lincs_fda <- lincs_filtered[lincs_filtered$pert_iname %in% overlap, ]

cat("LINCS profiles for overlapping drugs:", nrow(lincs_fda), "\n")
cat("Unique overlapping drugs in LINCS:", length(unique(lincs_fda$pert_iname)), "\n")
cat("Number of cell lines:", length(unique(lincs_fda$cell_id)), "\n")

cat("Cell lines:\n")
print(unique(lincs_fda$cell_id))
```

```{r}
compute_metrics <- function(drug_list, positives, K) {
  drug_list <- tolower(drug_list)
  
  topk <- drug_list[1:K]
  
  hits <- sum(topk %in% positives)
  
  precision <- hits / K
  recall <- hits / length(positives)
  
  c(Precision = precision, Recall = recall)
}
```

```{r}
labels = labels[labels$drug_name %in% overlap,]
labels$drug_name <- tolower(labels$drug_name)
positives <- labels$drug_name[labels$HDT_label == 1]
length(positives)
```

```{r}
library(dplyr)

rank_df <- lapply(names(method_lists), function(method) {
  
  drugs <- tolower(method_lists[[method]])
  
  data.frame(
    Method = method,
    drug_name = drugs,
    Rank = seq_along(drugs),
    signature_type = ifelse(grepl("aggr", method, ignore.case = TRUE),
                            "Aggregated signatures",
                            "Individual signatures")
  )
}) %>%
  bind_rows()

# keep only known positives
rank_df <- rank_df %>%
  filter(drug_name %in% positives)
rank_df$Score <- gsub("_.*", "", rank_df$Method)

```

```{r}
library(ggplot2)
library(dplyr)

# order Score by median rank (lower = better)
rank_df <- rank_df %>%
  mutate(Score = gsub("_.*", "", Method)) %>%
  group_by(Score) %>%
  mutate(median_rank = median(Rank, na.rm = TRUE)) %>%
  ungroup()

rank_df$Score <- reorder(rank_df$Score, rank_df$median_rank)

ggplot(rank_df, aes(x = Score, y = Rank)) +
  geom_boxplot(alpha = 0.7) +
  scale_y_reverse() +  # rank 1 at top
  coord_flip() +       # flip axes
  theme_minimal() +
  labs(
    title = "Distribution of ranks for known drugs in the top 20 predictions across methods",
    x = "Method",
    y = "Rank (Top = 1)"
  ) +
  theme(legend.position = "none")
```

```{r}
library(dplyr)
library(ggplot2)

rank_df_expanded <- rank_df %>%
  mutate(signature_type = ifelse(Method == "Integrative", signature_type, signature_type)) %>%
  # add "All" panel for Integrative
  bind_rows(
    rank_df %>%
      filter(Method == "Integrative") %>%
      mutate(signature_type = "All (Integrative)")
  )

# reorder signature_type by median rank (ascending)
rank_df_expanded <- rank_df_expanded %>%
  mutate(signature_type = factor(signature_type, 
                                 levels = names(sort(tapply(Rank, signature_type, median)))))

ggplot(rank_df_expanded, aes(x = signature_type, y = Rank, fill = signature_type)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +    
  scale_y_reverse() +   # rank 1 at top
  coord_flip() +        # horizontal boxes
  theme_minimal() +
  labs(
    title = "Distribution of ranks for known drugs in \n the top 20 predictions by signature type",
    x = "Signature type",
    y = "Rank (Top = 1)"
  ) +
  theme(legend.position = "none")
```
```{r}
library(dplyr)
library(ggplot2)

rank_df_expanded <- rank_df %>%
  # assign Technology for non-integrative rows
  mutate(
    Technology = case_when(
      grepl("microarray", Method, ignore.case = TRUE) ~ "Microarray",
      grepl("rnaseq|rna-seq", Method, ignore.case = TRUE) ~ "RNAseq",
      TRUE ~ NA_character_
    )
  ) %>%
  
  # keep only rows with a defined technology
  filter(!is.na(Technology)) %>%
  
  # duplicate Integrative rows into BOTH facets and relabel signature_type
  bind_rows(
    rank_df %>%
      filter(Method == "Integrative") %>%
      mutate(
        Technology = "Microarray",
        signature_type = "Combined (integrative)"
      ),
    
    rank_df %>%
      filter(Method == "Integrative") %>%
      mutate(
        Technology = "RNAseq",
        signature_type = "Combined (integrative)"
      )
  )

# reorder signature_type (now includes "Combined")
rank_df_expanded <- rank_df_expanded %>%
  mutate(signature_type = factor(
    signature_type,
    levels = names(sort(tapply(Rank, signature_type, median)))
  ))
pdf(here("figures/figureX/figureX_overall_ranks.pdf"),
    width = 6, height = 4)
ggplot(rank_df_expanded, aes(x = signature_type, y = Rank, fill = signature_type)) +
  geom_boxplot(alpha = 0.7, outlier.shape = NA) +
  scale_y_reverse() +
  coord_flip() +
  facet_wrap(~ Technology) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Distribution of ranks for known drugs in the top 20 predictions by signature type",
    x = "Signature type",
    y = "Rank (Top = 1)"
  ) +
  theme(legend.position = "none")
dev.off()
```

```{r}
# set the order of methods
rank_df_plot$Score <- factor(rank_df_plot$Score,
                             levels = rev(c("CMAP", "WCS", "NCS", "Tau", "Pearson", "Spearman", "Integrative")))

# calculate median rank for Integrative
median_integrative <- rank_df_plot %>%
  filter(Score == "Integrative") %>%
  summarise(med = median(Rank)) %>%
  pull(med)

pdf(here("figures/figureX/figureX_rank_distributions.pdf"), width = 10, height = 12)
# Plot
ggplot(rank_df_plot, aes(
  x = Score,
  y = Rank,
  fill = signature_type,
  alpha = Score == "Integrative"  # TRUE/FALSE
)) +
  geom_boxplot(color = "black") +
  # ddd horizontal line at median of Integrative (vertical after flip)
  geom_hline(yintercept = median_integrative, linetype = "dashed", color = "red", size = 0.5) +
  scale_y_reverse() +
  coord_flip() +
  facet_wrap(~DataType) +
  theme_minimal(base_size = 16) +
  scale_fill_manual(
    values = c(
      "Aggregated signatures" = "#0072B2",
      "Individual signatures" = "#D55E00"
    )
  ) +
  scale_alpha_manual(
    values = c(`TRUE` = 0, `FALSE` = 0.8),  # Integrative vs others
    guide = "none"  # hide alpha legend
  ) +
  labs(
    title = "Distribution of ranks for known drugs in the top 20 predictions across methods",
    x = "Method",
    y = "Rank (Top = 1)",
    fill = "Signature Type"
  ) +
  theme(
    legend.position = "right",
    strip.text = element_text(face = "bold")
  )
dev.off()
```

```{r}
rank_df <- rank_df %>%
  mutate(
    Score = gsub("_.*", "", Method),
    DataType = ifelse(grepl("RNAseq", Method), "RNAseq",
                      ifelse(grepl("microarray", Method), "Microarray", "Integrative")),
    Method_full = ifelse(DataType == "Integrative",
                         "Integrative",
                         paste(Score, DataType, sep = "_"))
  )
rank_df <- rank_df %>%
  group_by(Method_full) %>%
  mutate(median_rank = median(Rank, na.rm = TRUE)) %>%
  ungroup()

rank_df$Method_full <- reorder(rank_df$Method_full, rank_df$median_rank)
```

```{r}
library(ggplot2)
library(dplyr)

rank_df <- rank_df %>%
  mutate(fill_color = ifelse(Method == "Integrative", "black",
                             ifelse(signature_type == "Aggregated signatures", "#1f77b4",
                                    "#ff7f0e")))

ggplot(rank_df, aes(x = Method_full, y = Rank, fill = fill_color)) +
  geom_boxplot(
    alpha = 0.7,
    position = position_dodge(width = 0.8),
    outlier.shape = NA
  ) +
  scale_fill_identity() +  
  scale_y_reverse() +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Distribution of ranks for known drugs \n by method-signature-type",
    x = "Method",
    y = "Rank (Top = 1)"
  ) +
  theme(legend.position = "right")
```

```{r}
Ks <- c(5, 10, 15, 20)

results <- lapply(Ks, function(k) {
  res <- lapply(method_lists, function(dl) {
    compute_metrics(dl, positives, K = k)
  })
  
  df <- as.data.frame(do.call(rbind, res))
  df$Method <- names(method_lists)
  df$K <- k
  df
})

results_df <- do.call(rbind, results)
```

```{r}
library(dplyr)

results_df_plot <- results_df %>%
  bind_rows(
    results_df %>%
      filter(grepl("Integrative", Method)) %>%
      mutate(DataType = "RNAseq")
  )

pdf(here("figures/figureX/figureX_precision@K_technology.pdf"),
    width = 8, height = 6)
ggplot(results_df_plot, aes(x = K, y = Precision, group = Method)) +
  
  geom_line(
    data = subset(results_df_plot, !grepl("Integrative", Method)),
    aes(color = Score),
    linewidth = 1,
    alpha = 0.5
  ) +
  geom_point(
    data = subset(results_df_plot, !grepl("Integrative", Method)),
    aes(color = Score),
    size = 3,
    alpha = 0.5
  ) +
  
  geom_line(
    data = subset(results_df_plot, grepl("Integrative", Method)),
    aes(color = Score),
    linewidth = 1.5,
    alpha = 1
  ) +
  geom_point(
    data = subset(results_df_plot, grepl("Integrative", Method)),
    aes(color = Score),
    size = 3.5,
    alpha = 1
  ) +
  
  facet_wrap(~ DataType) +
  scale_color_manual(values = score_colors) +
  theme_minimal() +
  labs(
    title = "Precision@K across methods by technology",
    x = "K",
    y = "Precision",
    color = "Score"
  )
dev.off()

```

```{r}
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(tidyr)

# add signature_type and IsIntegrative
results_df <- results_df %>%
  mutate(
    DataType = ifelse(grepl("RNAseq", Method), "RNAseq", "Microarray"),
    signature_type = case_when(
      grepl("ggr", Method) ~ "Aggregated signature",
      grepl("Integrative", Method) ~ "Aggregated signature", # Integrative treated as solid
      TRUE ~ "Individual signature"
    ),
    Score = gsub("_.*", "", Method),
    IsIntegrative = grepl("Integrative", Method)
  )

# duplicate Integrative rows into all facets without adding a new facet row
integrative_rows <- results_df %>%
  filter(IsIntegrative) %>%
  select(-DataType, -signature_type) %>%  # remove to avoid duplication errors
  crossing(
    DataType = c("RNAseq", "Microarray"),
    signature_type = c("Aggregated signature", "Individual signature") # Integrative appears in both facets
  )

results_df_plot <- bind_rows(results_df, integrative_rows)

# Define colors
score_colors <- c(
  "Integrative" = "black",
  setNames(brewer.pal(length(unique(results_df$Score)), "Dark2"),
           unique(results_df$Score))
)

pdf(here("figures/figureX/figureX_precision@K_technology_signature.pdf"),
    width = 8, height = 6)
ggplot(results_df_plot, aes(x = K, y = Precision, group = Method)) +
  
  geom_line(
    data = subset(results_df_plot, !IsIntegrative),
    aes(color = Score, linetype = signature_type),
    linewidth = 1,
    alpha = 0.5
  ) +
  geom_point(
    data = subset(results_df_plot, !IsIntegrative),
    aes(color = Score),
    size = 3,
    alpha = 0.5
  ) +
  
  geom_line(
    data = subset(results_df_plot, IsIntegrative),
    aes(color = Score),
    linewidth = 1.5,
    linetype = "solid",
    alpha = 1
  ) +
  geom_point(
    data = subset(results_df_plot, IsIntegrative),
    aes(color = Score),
    size = 3.5,
    alpha = 1
  ) +
  
  facet_grid(signature_type ~ DataType) +
  
  scale_color_manual(values = score_colors) +
  scale_linetype_manual(values = c(
    "Individual signature" = "dashed",
    "Aggregated signature" = "solid"
  )) +
  
  theme_minimal() +
  labs(
    title = "Precision@K across methods by technology and signature type",
    x = "K",
    y = "Precision",
    color = "Score",
    linetype = "Signature Type"
  ) +
  theme(
    strip.text = element_text(face = "bold")
  )
dev.off()
```

```{r}
library(dplyr)
library(ggplot2)
library(RColorBrewer)

# ensure signature_type exists
results_df <- results_df %>%
  mutate(
    signature_type = case_when(
      grepl("ggr", Method) ~ "Aggregated signature",
      grepl("Integrative", Method) ~ "Aggregated signature", # Integrative treated as Aggregated
      TRUE ~ "Individual signature"
    ),
    Score = gsub("_.*", "", Method)
  )

# compute average Precision@K by Score, signature_type, and K
avg_df <- results_df %>%
  group_by(Score, signature_type, K) %>%
  summarise(AveragePrecision = mean(Precision, na.rm = TRUE), .groups = "drop")

score_colors <- c(
  "Integrative" = "black",
  setNames(RColorBrewer::brewer.pal(length(unique(results_df$Score)), "Dark2"),
           unique(results_df$Score))
)

pdf(here("figures/figureX/figureX_average_precision@K_technology_signature.pdf"),
    width = 7, height = 4)
ggplot(avg_df, aes(x = K, y = AveragePrecision, color = Score)) +
  geom_line(
    aes(linetype = signature_type, alpha = ifelse(Score == "Integrative", 1, 0.5)),
    linewidth = 1.2
  ) +
  geom_point(
    aes(alpha = ifelse(Score == "Integrative", 1, 0.5)),
    size = 3
  ) +
  scale_color_manual(values = score_colors) +
  scale_linetype_manual(values = c(
    "Individual signature" = "dashed",
    "Aggregated signature" = "solid"
  )) +
  scale_alpha_identity() +   # use the alpha values as is
  theme_minimal() +
  labs(
    title = "Average precision@K across microarray and RNAseq",
    x = "K",
    y = "Average precision",
    color = "Score",
    linetype = "Signature type"
  )
dev.off()
```

```{r}
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(tidyr)

# add signature_type and IsIntegrative
results_df <- results_df %>%
  mutate(
    DataType = ifelse(grepl("RNAseq", Method), "RNAseq", "Microarray"),
    signature_type = case_when(
      grepl("ggr", Method) ~ "Aggregated signature",
      grepl("Integrative", Method) ~ "Aggregated signature", # Integrative treated as solid
      TRUE ~ "Individual signature"
    ),
    Score = gsub("_.*", "", Method),
    IsIntegrative = grepl("Integrative", Method)
  )

# duplicate Integrative rows into all facets without adding a new facet row
integrative_rows <- results_df %>%
  filter(IsIntegrative) %>%
  select(-DataType, -signature_type) %>%  # remove to avoid duplication errors
  crossing(
    DataType = c("RNAseq", "Microarray"),
    signature_type = c("Aggregated signature", "Individual signature") # Integrative appears in both facets
  )

results_df_plot <- bind_rows(results_df, integrative_rows)

score_colors <- c(
  "Integrative" = "black",
  setNames(brewer.pal(length(unique(results_df$Score)), "Dark2"),
           unique(results_df$Score))
)
pdf(here("figures/figureX/figureX_recall@K_technology_signature.pdf"),
    width = 8, height = 6)
ggplot(results_df_plot, aes(x = K, y = Recall, group = Method)) +
  
  geom_line(
    data = subset(results_df_plot, !IsIntegrative),
    aes(color = Score, linetype = signature_type),
    linewidth = 1,
    alpha = 0.5
  ) +
  geom_point(
    data = subset(results_df_plot, !IsIntegrative),
    aes(color = Score),
    size = 3,
    alpha = 0.5
  ) +
  
  geom_line(
    data = subset(results_df_plot, IsIntegrative),
    aes(color = Score),
    linewidth = 1.5,
    linetype = "solid",
    alpha = 1
  ) +
  geom_point(
    data = subset(results_df_plot, IsIntegrative),
    aes(color = Score),
    size = 3.5,
    alpha = 1
  ) +
  
  facet_grid(signature_type ~ DataType) +
  
  scale_color_manual(values = score_colors) +
  scale_linetype_manual(values = c(
    "Individual signature" = "dashed",
    "Aggregated signature" = "solid"
  )) +
  
  theme_minimal() +
  labs(
    title = "Recall@K across methods by technology and signature type",
    x = "K",
    y = "Recall",
    color = "Score",
    linetype = "Signature type"
  ) +
  theme(
    strip.text = element_text(face = "bold")
  )
dev.off()
```

```{r}
library(dplyr)
library(ggplot2)
library(RColorBrewer)

# ensure signature_type exists
results_df <- results_df %>%
  mutate(
    signature_type = case_when(
      grepl("ggr", Method) ~ "Aggregated signature",
      grepl("Integrative", Method) ~ "Aggregated signature", # Integrative treated as Aggregated
      TRUE ~ "Individual signature"
    ),
    Score = gsub("_.*", "", Method)
  )

# compute average Precision@K by Score, signature_type, and K
avg_df_recall <- results_df %>%
  group_by(Score, signature_type, K) %>%
  summarise(AverageRecall = mean(Recall, na.rm = TRUE), .groups = "drop")

score_colors <- c(
  "Integrative" = "black",
  setNames(RColorBrewer::brewer.pal(length(unique(results_df$Score)), "Dark2"),
           unique(results_df$Score))
)

pdf(here("figures/figureX/figureX_average_recall@K_technology_signature.pdf"),
    width = 7, height = 4)
ggplot(avg_df_recall, aes(x = K, y = AverageRecall, color = Score)) +
  geom_line(
    aes(linetype = signature_type, alpha = ifelse(Score == "Integrative", 1, 0.5)),
    linewidth = 1.2
  ) +
  geom_point(
    aes(alpha = ifelse(Score == "Integrative", 1, 0.5)),
    size = 3
  ) +
  scale_color_manual(values = score_colors) +
  scale_linetype_manual(values = c(
    "Individual signature" = "dashed",
    "Aggregated signature" = "solid"
  )) +
  scale_alpha_identity() +   
  theme_minimal() +
  labs(
    title = "Average recall@K across microarray and RNAseq",
    x = "K",
    y = "Average recall",
    color = "Score",
    linetype = "Signature type"
  )
dev.off()
```