# Get background genes

# Last modified: 10/28/2025

# Author: Kewalin Samart

##### **Load libraries and sort scripts**

```{r}
library(readr)
library(here)
source(here("scripts/00_background_genes_PA_functions.R"))
```

##### **Get background genes from human Gene Ontology database**

```{r}
GO_genes_bg <- GO_genes()
head(GO_genes_bg)
```

##### **Get background genes from human KEGG pathway database**

```{r}
KEGG_genes_bg <- KEGG_genes()
head(KEGG_genes_bg)
```

##### **Get background genes from LINCS**

```{r}
LINCS_genes_bg <- bg_LINCS_genes()
head(LINCS_genes_bg)
```

##### **Get background genes from data**

```{r}
metadata_path <- here("data/v2/signatures/RNAseq_TB_signature_run_info_mixed.tsv")
data_path <- here("data/v2/DE_results/RNAseq/up")
extension <- "_up.tsv"
data_genes_bg <- bg_from_data(metadata_path = metadata_path, data_path = data_path, extension = extension)
head(data_genes_bg)
```

##### **Get background genes from any source above**

```{r}
tryCatch(
    {
        bg_genes_GO <- get_bg_genes(bg_source = "GO")
        cat("GO genes background size:", length(bg_genes_GO), "\n")

        bg_genes_KEGG <- get_bg_genes(bg_source = "KEGG")
        cat("KEGG genes background size:", length(bg_genes_KEGG), "\n")

        bg_genes_LINCS <- get_bg_genes(bg_source = "LINCS")
        cat("LINCS genes background size:", length(bg_genes_LINCS), "\n")

		cat("\nRUNNING ON INPUT DATA...")

        bg_genes_INPUT <- get_bg_genes(bg_source = "input data", metadata_path = metadata_path, data_path = data_path, extension = extension)
        cat("Input data genes background size:", length(bg_genes_INPUT), "\n")
    },
    error = function(e) {
        cat("Error: ", e$message, "\n")
    }
)
```

#### **Get combined background genes from microarray and RNAseq intersecting with genes annotated to GO ontology**

```{r define arguments}
metadata_path_rnaseq <- here("data/v2/signatures/RNAseq_TB_signature_run_info_mixed.tsv")
data_path_rnaseq <- here("data/v2/DE_results/RNAseq")

metadata_path_microarray <- here("data/v2/signatures/microarray_TB_signature_run_info_mixed.tsv")
data_path_microarray <- here("data/v2/DE_results/microarray")

GO_genes_vector <- GO_genes()
```

```{r get upregulated combined background}
bg_genes_combined_up <- get_combined_bg_genes(
  metadata_path_rnaseq, data_path_rnaseq,
  metadata_path_microarray, data_path_microarray,
  direction = "up",
  GO_genes_vector,
  combine = "union"
)
print(paste0("Number of upregulated combined background genes: ", length(bg_genes_combined_up)))
```

```{r get downregulated combined background}
bg_genes_combined_dn <- get_combined_bg_genes(
  metadata_path_rnaseq, data_path_rnaseq,
  metadata_path_microarray, data_path_microarray,
  direction = "dn",
  GO_genes_vector,
  combine = "union"
)
print(paste0("Number of downregulated combined background genes: ", length(bg_genes_combined_dn)))
```