# Compute aggregated signatures from individual signatures
# for pathway enrichment analyses (all genes present across datasets are background genes)
# these are done separately by technology-direction
# Last modified: 08/27/25
# Author: Kewalin Samart

##### Load libraries and sort scripts
```{r}
library(readr)
library(here)
source(here("scripts/00_background_genes_PA_functions.R"))
source(here("scripts/01_DE_signature_aggregation_functions.R"))
```

##### Get background genes
```{r}
# define input paths
metadata_path_rnaseq <- here("data/v2/signatures/RNAseq_TB_signature_run_info.tsv")
data_path_rnaseq <- here("data/v2/DE_results/RNAseq")

metadata_path_microarray <- here("data/v2/signatures/microarray_TB_signature_run_info.tsv")
data_path_microarray <- here("data/v2/DE_results/microarray")
GO_genes_vector <- GO_genes()

# get upregulated combined background
bg_genes_combined_up <- get_combined_bg_genes(
  metadata_path_rnaseq, data_path_rnaseq,
  metadata_path_microarray, data_path_microarray,
  direction = "up",
  GO_genes_vector,
  combine = "union"
)

# get downregulated combined background
bg_genes_combined_dn <- get_combined_bg_genes(
  metadata_path_rnaseq, data_path_rnaseq,
  metadata_path_microarray, data_path_microarray,
  direction = "dn",
  GO_genes_vector,
  combine = "union"
)
```

##### Initialize arguments
```{r}
# example argument
metadata_path = here("data/v2/signatures/RNAseq_TB_signature_run_info.tsv")
data_path = here("data/v2/DE_results/RNAseq")
output_dir = here("data/v2/signatures/RNAseq/aggregated_DE_signatures")
direction = "dn"
threshold_pct <- 0.90
```

##### Compute membership matrix
```{r}
gene_membership_matrix = compute_membership_matrix(metadata_path = metadata_path, data_path = data_path, 
                          direction = direction, bg_genes = bg_genes_combined_dn, 
                          output_dir = output_dir,save_result=TRUE)
```

##### Compute jaccard matrix
```{r}
jaccard_matrix = compute_jaccard_matrix(metadata_path = metadata_path, data_path = data_path, 
                       direction = direction, output_dir = output_dir, 
                       extra_arg = extra_arg, save_result=TRUE) 
```

##### Compute aggregated signature
```{r}
aggregate_signatures(gene_membership_matrix, jaccard_matrix, output_dir, direction, threshold_pct = threshold_pct, save_result=TRUE)
```
