# Compute aggregated signatures from individual signatures
# for pathway enrichment analyses (all genes present across datasets are background genes; combined microarray+RNAseq per direction)
# these are done separately by technology-direction
# the differentially expressed genes are passed to pathway analysis 
# individual signatures: adjusted p-val < 0.05; not restricted to only the landmark genes
# aggregated signatures: all genes (not just the landmark genes) with > 90% pct of all the aggregated gene scores

##### Load libraries and sort scripts
```{r}
library(readr)
library(here)
library(tidyverse)
library(clusterProfiler)
library(org.Hs.eg.db)
library(GO.db)
library(jsonlite)
library(tidyr)
library(qvalue)
source(here("scripts/00_background_genes_PA_functions.R"))
```

##### Get background genes
```{r}
# define input paths
metadata_path_rnaseq <- here("data/signatures/RNAseq_TB_signature_run_info.tsv")
data_path_rnaseq <- here("data/DE_results/RNAseq")

metadata_path_microarray <- here("data/signatures/microarray_TB_signature_run_info.tsv")
data_path_microarray <- here("data/DE_results/microarray")
GO_genes_vector <- GO_genes()

# get upregulated combined background
bg_genes_combined_up <- get_combined_bg_genes(
  metadata_path_rnaseq, data_path_rnaseq,
  metadata_path_microarray, data_path_microarray,
  direction = "up",
  GO_genes_vector,
  combine = "union"
)

# get downregulated combined background
bg_genes_combined_dn <- get_combined_bg_genes(
  metadata_path_rnaseq, data_path_rnaseq,
  metadata_path_microarray, data_path_microarray,
  direction = "dn",
  GO_genes_vector,
  combine = "union"
)
```

##### Pathway enrichment analysis 

##### Get all GO:BP terms
```{r}
library(clusterProfiler)
library(org.Hs.eg.db)

# set background genes
bg_genes <- union(bg_genes_combined_up, bg_genes_combined_dn)

# parameters
ont <- "BP"
minGSSize <- 5
maxGSSize <- 200

# extract GO data as used by enrichGO()
godata <- clusterProfiler:::get_GO_data(
  OrgDb = org.Hs.eg.db,
  ont = ont,
  keytype = "ENTREZID"
)

# load GO data (all genes from OrgDb)
godata <- clusterProfiler:::get_GO_data(
  OrgDb = org.Hs.eg.db,
  ont = "BP",
  keytype = "ENTREZID"
)

# filter gene sets to include only genes in universe
go_list <- godata$PATHID2EXTID

# build data frame of term descriptions
go_terms <- data.frame(
  GO_ID = names(go_list),
  Description = Term(GOTERM[names(go_list)]),
  GeneSetSize = sapply(go_list, function(x) length(unique(intersect(x, bg_genes)))),
  stringsAsFactors = FALSE
)
```

###### Calculate enrichment metrics
```{r}
#------------------------------------------------------------------------------
# calculate_asinh_enrichment
#
# Description:
#   Computes enrichment scores for pathways by comparing observed vs expected
#   overlaps between a signature gene list and pathway gene sets. Adds two
#   metrics: 
#     - log(observed / expected)
#     - asinh(observed / expected) for stabilized values
#
# Arguments:
#   enrichment_df        Data frame with pathway enrichment results.
#                        Must include columns:
#                          - ID: pathway identifier
#                          - GeneRatio: string of the form "k/n"
#   signature_gene_list  Vector of genes in the signature set (S).
#   bg_genes             Vector of all background genes (U).
#   pathway_df   Data frame with pathway sizes.
#                        Must include columns:
#                          - ID: pathway identifier
#                          - pathway_genes: all propagated genes to the pathway
#                          - pathway_size: pathway size (|P|).
#
# Returns:
#   The input data frame with added columns:
#     - bp_gene_length: size of the pathway
#     - log_observed_expected: log ratio of observed vs expected overlap
#     - asinh_observed_expected: asinh-transformed observed/expected ratio
#
#------------------------------------------------------------------------------

calculate_asinh_enrichment <- function(signature_gene_list, bg_genes, pathway_df) {
  
  # ensure pathway_genes column is list-type (if it's character, split it)
  if (is.character(pathway_df$pathway_genes)) {
    pathway_df$pathway_genes <- lapply(
      pathway_df$pathway_genes,
      function(x) {
        x <- gsub("^c\\(|\\)$", "", x)  # remove c( and )
        x <- gsub('\"', "", x)          # remove double quotes
        x <- strsplit(x, split = ",\\s*")[[1]]  # split on commas
        x
      }
    )
  }

  # calculate observed as actual intersection between pathway genes and signature genes
  pathway_df$observed_val <- sapply(
    pathway_df$pathway_genes,
    function(genes) length(intersect(genes, signature_gene_list))
  )
  
  pathway_df$pathway_genes <- sapply(pathway_df$pathway_genes, function(x) {
    # remove "c(", ")", and double quotes
    x <- gsub('^c\\(|\\)$', '', x)       # remove leading c( and trailing )
    x <- gsub('""', '', x)               # remove double double-quotes
    x <- gsub('"', '', x)                # remove any leftover single double-quotes
    x <- gsub('\\s+', '', x)             # remove whitespace if needed
    return(x)                            # already comma-separated
  })
  
  # expected = (|P| / |U|) * |S|
  pathway_df$expected_val <- (pathway_df$pathway_size / length(bg_genes)) * length(signature_gene_list)
  
  # calculate log and asinh enrichment
  pathway_df$log_observed_expected <- log(pathway_df$observed_val / pathway_df$expected_val)
  pathway_df$asinh_observed_expected <- asinh(pathway_df$observed_val / pathway_df$expected_val)
  
  # hypergeometric p-values
  pathway_df$pvalue <- mapply(
  function(k, pathway_genes_i) {

    N <- length(bg_genes)                           # total universe
    n <- length(signature_gene_list)                # signature size
    M <- length(intersect(pathway_genes_i, bg_genes))  # pathway size in universe

    phyper(k - 1, M, N - M, n, lower.tail = FALSE)
  },
  k = pathway_df$observed_val,
  pathway_genes_i = pathway_df$pathway_genes       # row-by-row list input
)

  
  return(pathway_df)
}

compute_multiple_testing <- function(pathway_df) {
  # BH adjusted p-values
  pathway_df$p.adjust <- p.adjust(pathway_df$pvalue, method = "BH")

  # Storey q-values (same meaning as clusterProfiler qvalue column)
  if (!requireNamespace("qvalue", quietly = TRUE)) {
    stop("Please install the qvalue package: install.packages('qvalue') or BiocManager::install('qvalue')")
  }

  pathway_df$qvalue <- qvalue::qvalue(pathway_df$pvalue)$qvalues

  return(pathway_df)
}

```

```{r set up pathway df}
json_propagated_genes <- fromJSON(here("data/metadata/propagated_annotations.json"))
# convert json_propagated_genes into a dataframe
pathway_df <- data.frame(
  ID = names(json_propagated_genes),
  pathway_genes = I(unname(json_propagated_genes)),
  stringsAsFactors = FALSE
)
# add pathway_size column
pathway_df$pathway_size <- sapply(pathway_df$pathway_genes, length)
# only select pathways with minimum gene set size of 5 genes and maximum of 200 genes
pathway_df <- pathway_df[pathway_df$pathway_size >= 5 | pathway_df$pathway_size <= 200, ]
pathway_df  <- merge(pathway_df, go_terms, by.x = "ID", by.y = "GO_ID", all.x = TRUE)[c("Description","ID","pathway_genes","pathway_size")]
```

###### Individual signatures 
```{r}
technologies <- c("microarray", "RNAseq")
directions <- c("up", "dn")

for (technology in technologies) {
  
  data_path <- here(paste0("data/DE_results/",technology))
  data_to_read <- read_tsv(here(paste0("data/signatures/", technology, "_TB_signature_run_info.tsv")))
  data_to_read = data_to_read[data_to_read$signature == 1,]

  for (direction in directions) {

    # set output directory
    dirname <- here(paste0("data/", technology, "_pathways/", direction))
    if (!dir.exists(dirname)) {
      dir.create(dirname, recursive = TRUE)
    }

    # read gene files from up/dn folder
    dirname_read <- file.path(data_path, direction)
    filenames <- list.files(path = dirname_read, pattern = paste0(direction, ".tsv"))

    # define background genes
    if(direction == "up"){
      bg_genes_ <- bg_genes_combined_up
    }else if(direction == "dn"){
      bg_genes_ <- bg_genes_combined_dn
    }
    
    # restrict to GO-annotated genes
    bg_genes <- intersect(bg_genes_, GO_genes())  

    # iterate through each file
    itr <- 1
    for (i in 1:nrow(data_to_read)) {
      
      print(paste0("Running ", technology, " ", direction, " iteration: ", itr))
      itr <- itr + 1
      
      signature_name = data_to_read$signature_name[i]
      file = paste0(signature_name,"_",direction,".tsv")
      genes_df <- read_tsv(file.path(dirname_read, file))
      # only use differentially expressed genes with adjusted p-val < 0.05
      genes_df_filtered <- genes_df[genes_df$adj.P.Val < 0.05,]

      # use gene IDs as input for ORA
      gene_vector <- genes_df_filtered$log2FoldChange
      names(gene_vector) <- as.character(genes_df_filtered$GeneID)

      gene_list <- names(gene_vector)

      # run ORA
      print(paste0("Getting enriched GO pathways for ", file))

      BP_terms <- go_terms$GO_ID
      
      # only care about biological processes (BP)
      pathway_df_BP <- pathway_df[pathway_df$ID %in% BP_terms,]
      
      # REPLACED MANUAL CALCULATION WITH FUNCTION CALL
      enrichGO_res <- calculate_asinh_enrichment(
        signature_gene_list = gene_list,
        bg_genes = bg_genes,
        pathway_df = pathway_df_BP
      )
      # multiple-testing correction: add qvalue
      enrichGO_res_corrected <- compute_multiple_testing(enrichGO_res)
        
      # check if the universal bp terms are valid
      print("All BP terms are valid?")
      print(unique(enrichGO_res_corrected$ID %in% go_terms$GO_ID))
        
      # save results
      base_filename <- sub(paste0("\\.", direction, "\\.tsv$"), "", file)
      output_file <- file.path(dirname, paste0("allterms_GO_ORA_BGcorrected_", base_filename))

      write_tsv(enrichGO_res_corrected, file = output_file)
    }
  }
}
```

###### Aggregated signatures
```{r all signatures}
technologies <- c("microarray", "RNAseq")
directions <- c("up","dn")
for(technology in technologies){
  for(direction in directions){
    # set output directory
    dirname <- here(paste0("data/", technology, "_pathways/", direction))
    if (!dir.exists(dirname)) {
      dir.create(dirname, recursive = TRUE)
    }
    
    aggr_gene_list <- as.character(read_tsv(here(paste0("data/DE_results/",technology,"/aggregated_DE_signatures/",direction,"_aggregated_signature.tsv")))$GeneID)
    
    # define background genes
    if(direction == "up"){
      bg_genes <- bg_genes_combined_up
    }else if(direction == "dn"){
      bg_genes <- bg_genes_combined_dn
    }
        
    # run ORA
    print(paste0("Getting enriched GO pathways for data/DE_results/",technology,"/aggregated_DE_signatures/",direction,"_aggregated_signature.tsv"))

    BP_terms <- go_terms$GO_ID
      
    # only care about biological processes (BP)
    pathway_df_BP <- pathway_df[pathway_df$ID %in% BP_terms,]
      
    # REPLACED MANUAL CALCULATION WITH FUNCTION CALL
    enrichGO_res_aggr <- calculate_asinh_enrichment(
      signature_gene_list = aggr_gene_list,
      bg_genes = bg_genes,
      pathway_df = pathway_df_BP
    )
      # multiple-testing correction: add qvalue
    enrichGO_res_aggr_corrected <- compute_multiple_testing(enrichGO_res_aggr)
        
    # check if the universal bp terms are valid
    print("All BP terms are valid?")
    print(unique(enrichGO_res_aggr_corrected$ID %in% go_terms$GO_ID))
        
    # save results
    output_file <- file.path(dirname, paste0("allterms_GO_ORA_BGcorrected_",direction,"_aggregated_signature.tsv"))

    write_tsv(enrichGO_res_aggr_corrected, file = output_file)
  }
}
```

```{r microarray and RNAseq origin-only signatures}
technologies <- c("microarray","RNAseq")
directions <- c("up","dn")
tissue_origins <- c("circulating")
for(technology in technologies){
  for(direction in directions){
    for(tissue_origin in tissue_origins){
      # set output directory
      dirname <- here(paste0("data/", technology, "_pathways/", direction))
      if (!dir.exists(dirname)) {
        dir.create(dirname, recursive = TRUE)
      }
      originaggr_gene_list <- as.character(read_tsv(here(paste0("data/DE_results/",technology,"/",tissue_origin,"_aggregated_DE_signatures/",direction,"_aggregated_signature.tsv")))$GeneID)
    
      # define background genes
      if(direction == "up"){
        bg_genes <- bg_genes_combined_up
      }else if(direction == "dn"){
        bg_genes <- bg_genes_combined_dn
      }
        
      # run ORA
      print(paste0("Getting enriched GO pathways for data/DE_results/",technology,"/",tissue_origin,"_aggregated_DE_signatures/",direction,"_aggregated_signature.tsv"))

      BP_terms <- go_terms$GO_ID
      
      # only care about biological processes (BP)
      pathway_df_BP <- pathway_df[pathway_df$ID %in% BP_terms,]
      
      # REPLACED MANUAL CALCULATION WITH FUNCTION CALL
      enrichGO_res_originaggr <- calculate_asinh_enrichment(
        signature_gene_list = originaggr_gene_list,
        bg_genes = bg_genes,
        pathway_df = pathway_df_BP
      )
      # multiple-testing correction: add qvalue
      enrichGO_res_originaggr_corrected <- compute_multiple_testing(enrichGO_res_originaggr)
        
      # check if the universal bp terms are valid
      print("All BP terms are valid?")
      print(unique(enrichGO_res_originaggr_corrected$ID %in% go_terms$GO_ID))
        
      # save results
      output_file <- file.path(dirname, paste0("allterms_GO_ORA_BGcorrected_",direction,"_",tissue_origin,"_aggregated_signature.tsv"))

      write_tsv(enrichGO_res_originaggr_corrected, file = output_file)
    } 
  }
}
```

