# Compute aggregated signatures from individual signatures
# for pathway enrichment analyses (all genes present across datasets are background genes)
# these are done separately by technology-direction
# Last modified: 08/27/25
# Author: Kewalin Samart

##### Load libraries and sort scripts
```{r}
library(readr)
library(here)
library(tidyverse)
library(clusterProfiler)
library(org.Hs.eg.db)
library(GO.db)
source(here("scripts/00_background_genes_PA_functions.R"))
```

##### Get background genes
```{r}
# define input paths
metadata_path_rnaseq <- here("data/v2/signatures/RNAseq_TB_signature_run_info.tsv")
data_path_rnaseq <- here("data/v2/DE_results/RNAseq")

metadata_path_microarray <- here("data/v2/signatures/microarray_TB_signature_run_info.tsv")
data_path_microarray <- here("data/v2/DE_results/microarray")
GO_genes_vector <- GO_genes()

# get upregulated combined background
bg_genes_combined_up <- get_combined_bg_genes(
  metadata_path_rnaseq, data_path_rnaseq,
  metadata_path_microarray, data_path_microarray,
  direction = "up",
  GO_genes_vector,
  combine = "union"
)

# get downregulated combined background
bg_genes_combined_dn <- get_combined_bg_genes(
  metadata_path_rnaseq, data_path_rnaseq,
  metadata_path_microarray, data_path_microarray,
  direction = "dn",
  GO_genes_vector,
  combine = "union"
)
```

##### Construct pathway-gene annotation dictionary
```{r}
# read in the pathway-gene annotation file
# we filter to class=BP and must have one of the following evidence codes: EXP, IDA, IPI, IMP, IGI, TAS, IC (don't remember what they stand for)
term_gene_annotation = read_tsv(here("data/metadata/direct-annotations__Homo_sapiens__Entrez.tsv"))
bp_gene_mapping_df = term_gene_annotation %>% filter(Class == "BP") %>%
  filter(Evidence %in% c("EXP", "IDA", "IPI", "IMP", "IGI", "TAS", "IC"))
bp_gene_mapping <- split(bp_gene_mapping_df$EntrezID, bp_gene_mapping_df$GOID)
bp_gene_length <- sapply(bp_gene_mapping, length)
bp_gene_length_df <- as.data.frame(bp_gene_length)
bp_gene_length_df$ID <- row.names(bp_gene_length_df)
```

##### Pathway enrichment analysis 
```{r}
technologies <- c("microarray", "RNAseq")
directions <- c("up", "dn")

for (technology in technologies) {

  metadata_path <- here(paste0("data/v2/signatures/", technology, "_TB_signature_run_info.tsv"))
  data_path <- here(paste0("data/v2/DE_results/",technology))

  for (direction in directions) {

    # set output directory
    dirname <- here(paste0("data/v2/", technology, "_pathways/", direction))
    if (!dir.exists(dirname)) {
      dir.create(dirname, recursive = TRUE)
    }

    # read gene files from up/dn folder
    dirname_read <- file.path(data_path, direction)
    filenames <- list.files(path = dirname_read, pattern = paste0(direction, ".tsv"))

    # define background genes
    if(direction == "up"){
      bg_genes_ <- bg_genes_combined_up
    }else if(direction == "dn"){
      bg_genes_ <- bg_genes_combined_dn
    }
    
    # restrict to GO-annotated genes
    bg_genes <- intersect(bg_genes_, GO_genes())  

    # iterate through each file
    itr <- 1
    for (file in filenames) {

      print(paste0("Running ", technology, " ", direction, " iteration: ", itr))
      itr <- itr + 1

      genes_df <- read_tsv(file.path(dirname_read, file))
      genes_df_filtered <- genes_df[genes_df$adj.P.Val < 0.05,]

      # use gene IDs as input for ORA
      gene_vector <- genes_df_filtered$log2FoldChange
      names(gene_vector) <- as.character(genes_df_filtered$GeneID)

      gene_list <- names(gene_vector)

      # run ORA
      print(paste0("Getting enriched GO pathways for ", file))

      enrichGO_res <- enrichGO(gene = gene_list,
                               OrgDb = org.Hs.eg.db,
                               readable = TRUE,
                               ont = "BP",
                               universe = bg_genes,
                               pvalueCutoff = 0.05,
                               qvalueCutoff = 0.1,
                               minGSSize = 5,
                               maxGSSize = 200)

      enrichGO_res <- as.data.frame(enrichGO_res)
      
      if (!is.null(enrichGO_res) && nrow(enrichGO_res) > 0) {
      
        enrichGO_res_corrected = merge(enrichGO_res, bp_gene_length_df, by="ID", all.x = TRUE)
      
        # make sure GeneRatio column is numeric
        enrichGO_res_corrected$GeneRatio <- sapply(
          as.character(enrichGO_res_corrected$GeneRatio),
          function(x) {
            parts <- as.numeric(unlist(strsplit(x, "/")))
            parts[1] / parts[2]
          }
        )
      
        # calculate log(observed / expected)
        enrichGO_res_corrected$log_observed_expected = log(enrichGO_res_corrected$GeneRatio * length(gene_list) / enrichGO_res_corrected$bp_gene_length)
      
        # save results
        base_filename <- sub(paste0("\\.", direction, "\\.tsv$"), "", file)
        output_file <- file.path(dirname, paste0("GO_ORA_BGcorrected_", base_filename))

        write_tsv(enrichGO_res_corrected, file = output_file)
      }else{
        message("No enriched GO terms, skipping")
      }
    }
  }
}

```


