# Compute disease-drug connectivity scores via signatureSearch

#### Last modified: 06/17/25

#### Author: Kewalin Samart

```{r import functions/libraries}
library(signatureSearch)
library(here)
library(readr)

source(here("scripts/01_signature_landmark_prep_functions.R"))
source(here("scripts/02_signatureSearch_connectivity_scores_functions.R"))
```

### Functions to retrieve drug database from signatureSearchData

```{r set up CMap drug database}
cmap_path = setup_cmap1db() # this can take a few minutes
print(cmap_path)
```

```{r set up LINCS drug database}
lincs_path = setup_lincsdb() # this can take a few minutes (much longer than CMap)
print(lincs_path)
```

### Functions to obtain precomputed disease signatures

#### Individual signatures

#### up and down regulated gene sets for CMAP 1.0 and LINCS methods (enrichment-based methods)

```{r obtain up/dn genes from up/dn signatures}
# get upregulated gene list
up_TB_signature_path = here("data/v2/signatures/RNAseq/up/GSE64179_MTB_control_blood_dendritic_cells_up.tsv")
up_genes = get_updn_signature(up_TB_signature_path, L1000 = TRUE)
print("Example up genes:")
print(head(up_genes))

# get downregulated gene list
dn_TB_signature_path = here("data/v2/signatures/RNAseq/dn/GSE64179_MTB_control_blood_dendritic_cells_dn.tsv")
dn_genes = get_updn_signature(dn_TB_signature_path, L1000 = TRUE)
print("Example down genes:")
print(head(dn_genes))
```

#### full signatures (up+dn) for XCor and XSpe (correlation-based methods)

```{r combine up and dn signatures into a full signature}
up_TB_signature = read_tsv(here("data/v2/signatures/RNAseq/up/GSE64179_MTB_control_blood_dendritic_cells_up.tsv"))
dn_TB_signature = read_tsv(here("data/v2/signatures/RNAseq/dn/GSE64179_MTB_control_blood_dendritic_cells_dn.tsv"))
full_signature = combine_signatures(up_signature = up_TB_signature, 
                                    dn_signature = dn_TB_signature, 
                                    L1000 = TRUE)
head(full_signature)
```

```{r preping full signatures for correlation-based methods}
full_TB_signature_path = here("data/v2/signatures/RNAseq/full/GSE64179_MTB_control_blood_dendritic_cells_full.tsv")
full_TB_signature = get_full_signature(full_sig_path = full_TB_signature_path,
                    L1000 = TRUE)
head(full_TB_signature)
```

#### Aggregated signatures

#### up and down regulated gene sets for CMAP 1.0 and LINCS methods (enrichment-based methods)

```{r obtain up/dn genes from up/dn signatures}
# get upregulated gene list
up_aggr_TB_signature_path = here("data/v2/signatures/RNAseq/aggregated_signatures/up_aggregated_signature.tsv")
up_aggr_genes = as.character(read_tsv(up_aggr_TB_signature_path)$GeneID)
print("Example up genes:")
print(head(up_aggr_genes))

# get downregulated gene list
dn_aggr_TB_signature_path = here("data/v2/signatures/RNAseq/aggregated_signatures/dn_aggregated_signature.tsv")
dn_aggr_genes = as.character(read_tsv(dn_aggr_TB_signature_path)$GeneID)
print("Example down genes:")
print(head(dn_aggr_genes))
```

#### full signatures (up+dn) for XCor and XSpe (correlation-based methods)

```{r combine up and dn signatures into a full signature}
up_aggr_TB_signature = read_tsv(here("data/v2/signatures/RNAseq/aggregated_signatures/up_aggregated_signature.tsv"))
dn_aggr_TB_signature = read_tsv(here("data/v2/signatures/RNAseq/aggregated_signatures/dn_aggregated_signature.tsv"))
combined_aggr_TB_signature <- rbind(up_aggr_TB_signature, dn_aggr_TB_signature)
combined_aggr_TB_signature <- combined_aggr_TB_signature[order(-combined_aggr_TB_signature$aggregated_GeneScores), ]

#combined_aggr_TB_signature$GeneID <- NULL
full_aggr_TB_signature <- as.matrix(combined_aggr_TB_signature[, !colnames(combined_aggr_TB_signature) %in% "GeneID"])
rownames(full_aggr_TB_signature) <- as.character(combined_aggr_TB_signature$GeneID)

head(full_aggr_TB_signature)
```

### Functions to calculate disease-drug connectivity scores

### Individual signatures

#### Enrichment-based methods: CMAP1.0 (CS), LINCS (WCS, NCS, Tau)

```{r CMAP 1.0}
cmap1.0_res = compute_CMap1_scores(up_genes = up_genes, 
                                   dn_genes = dn_genes, 
                                   db_path = lincs_path)
head(cmap1.0_res)
print(dim(cmap1.0_res))
```

```{r LINCS}
lincs_res = compute_CMap2lincs_scores(up_genes = up_genes, 
                                   dn_genes = dn_genes, 
                                   db_path = lincs_path)
head(lincs_res)
print(dim(lincs_res))
```

#### Correlation-based methods: XCor (extreme Pearson), XSpe (extreme Spearman)

```{r XCor}
xcor_res = compute_Cor_based_scores(full_signature_matrix = full_TB_signature, 
                                    score_method = 'Cor_pearson', 
                                    db_path = lincs_path)
head(xcor_res)
print(dim(xcor_res))
```

```{r XSpe}
xspe_res = compute_Cor_based_scores(full_signature_matrix = full_TB_signature, 
                                    score_method = 'Cor_spearman', 
                                    db_path = lincs_path)
head(xspe_res)
print(dim(xspe_res))
```

### Aggregated signatures

#### Enrichment-based methods: CMAP1.0 (CS), LINCS (WCS, NCS, Tau)

```{r CMAP 1.0}
cmap1.0_aggr_res = compute_CMap1_scores(up_genes = up_aggr_genes, 
                                   dn_genes = dn_aggr_genes, 
                                   db_path = lincs_path)
head(cmap1.0_aggr_res)
print(dim(cmap1.0_aggr_res))
```

```{r LINCS}
lincs_aggr_res = compute_CMap2lincs_scores(up_genes = up_aggr_genes, 
                                   dn_genes = dn_aggr_genes, 
                                   db_path = lincs_path)
head(lincs_aggr_res)
print(dim(lincs_aggr_res))
```

#### Correlation-based methods: XCor (extreme Pearson), XSpe (extreme Spearman)

```{r XCor}
xcor_aggr_res = compute_Cor_based_scores(full_signature_matrix = full_aggr_TB_signature, 
                                    score_method = 'Cor_pearson', 
                                    db_path = lincs_path)
head(xcor_aggr_res)
print(dim(xcor_aggr_res))
```

```{r XSpe}
xspe_aggr_res = compute_Cor_based_scores(full_signature_matrix = full_aggr_TB_signature, 
                                    score_method = 'Cor_spearman', 
                                    db_path = lincs_path)
head(xspe_aggr_res)
print(dim(xspe_aggr_res))
```
