# Functions to summarize drug results prioritized by multiple methods

## method-wise summarization

```{r}
library(here)
source(here("scripts/03_summarize_drugs_methodswise_functions.R"))
```

## Individual signatures
### RNAseq

1.  Select drugs reversing at least three individual disease signatures at the 80th percentile reversal threshold for all the six connectivity scores

```{r}
metadata_path <- "data/signatures/RNAseq_TB_signature_run_info.tsv"
all_signi_drugs_df <- create_drug_method_table(metadata_path, stats = "min", score_percentile = 0.80, n = 3)
if (!dir.exists(here("results/RNAseq/03_methodwise"))) {
      dir.create(here("results/RNAseq/03_methodwise"), recursive = TRUE)
    }
write_tsv(as.data.frame(all_signi_drugs_df$count_df), file = here("results/RNAseq/03_methodwise/RNAseq_drug_score_occurrence.tsv"))
```

2.  Summarize drugs by method categories: only select drugs that come up in two out of the three method categories (CMAP, LINCS, Cor)

```{r}
methodwise_results = summarize_drugs_bymethods(all_signi_drugs_df$binary_df, technology = "RNAseq", dirname = "results")
```

### microarray

1.  Select drugs reversing at least three individual disease signatures at the 80th percentile reversal threshold for all the six connectivity scores

```{r}
metadata_path <- "data/signatures/microarray_TB_signature_run_info.tsv"
all_signi_drugs_df <- create_drug_method_table(metadata_path, stats = "min", score_percentile = 0.8, n = 3)
if (!dir.exists(here("results/microarray/03_methodwise"))) {
      dir.create(here("results/microarray/03_methodwise"), recursive = TRUE)
    }
write_tsv(as.data.frame(all_signi_drugs_df$count_df), file = here("results/microarray/03_methodwise/microarray_drug_score_occurrence.tsv"))
```

2.  Summarize drugs by method categories: only select drugs that come up in two out of the three method categories (CMAP, LINCS, Cor)

```{r}
methodwise_results = summarize_drugs_bymethods(all_signi_drugs_df$binary_df, technology = "microarray", dirname = "results")
methodwise_results 
```

## Aggregated signatures
### RNAseq
1.  Obtain drugs passing 90% reversal for all the six connectivity scores
```{r}
technologies = c("microarray", "RNAseq")
scores = c("CMAP", "WCS", "NCS", "Tau", "Cor_spearman", "Cor_pearson")
values = "neg"
tail = "left"
score_percentile = 0.9
base_dir = "results"
```

```{r}
aggr_res = extract_aggrSig_significant_drugs(technologies = technologies,
                                             scores = scores,
                                             values = values,
                                             tail = tail,
                                             score_percentile = score_percentile,
                                             base_dir = base_dir)
```

```{r}
aggr_methodwise_res = summarize_methodwise_aggrSig(aggr_res)
head(aggr_methodwise_res)
```
