# Functions to summarize drug results prioritized by multiple methods

## method-wise summarization

#### Last modified: 06/17/25

#### Author: Kewalin Samart

```{r}
library(here)
source(here("scripts/03_summarize_drugs_methodswise_functions.R"))
```

### RNAseq

1.  Obtain drugs within top 3 occurrences passing 90% reversal for all the six connectivity scores

```{r}
metadata_path <- "data/v2/signatures/RNAseq_TB_signature_run_info.tsv"
all_signi_drugs_df <- create_drug_method_table(metadata_path, stats = "min", score_percentile = 0.9, n = 3)
all_signi_drugs_df
```

2.  Summarize drugs by method categories: only select drugs that come up in two out of the three method categories (CMAP, LINCS, Cor)

```{r}
methodwise_results = summarize_drugs_bymethods(all_signi_drugs_df, technology = "RNAseq", dirname = "results")
methodwise_results
```

### microarray

1.  obtain drugs within top 3 occurrences passing 90% reversal for all the six connectivity scores

```{r}
metadata_path <- "data/v2/signatures/microarray_TB_signature_run_info.tsv"
all_signi_drugs_df <- create_drug_method_table(metadata_path, stats = "min", score_percentile = 0.9, n = 3)
all_signi_drugs_df
```

2.  Summarize drugs by method categories: only select drugs that come up in two out of the three method categories (CMAP, LINCS, Cor)

```{r}
methodwise_results = summarize_drugs_bymethods(all_signi_drugs_df, technology = "microarray", dirname = "results")
methodwise_results 
```
