# High confidence drug predictions
This script processes individual and aggregated drug prediction results to identify high-confidence drug candidates based on their rank scores. calls Functions from `05_high_confidence_drug_prediction.R` to perform the following steps:

1. Load necessary libraries and source functions `read_pipeline_results()`
2. Load individual and aggregated drug prediction results `process_pipeline()`
3. Calculate mean rank scores for individual predictions `process_pipeline()`
4. Convert mean rank scores to z-scores `convert_to_zscore()`
5. Merge individual and aggregated results `get_high_confidence_predictions()`
6. Calculate combined z-scores for high-confidence predictions `get_high_confidence_predictions()`

## High Confidence Drug Predictions

#### Last modified: 12/18/25 

#### Author: Ling Thang


```{r setup}
library(here)
source(here("scripts/05_high_confidence_drug_prediction.R"))
```

##### Define paths to the rank-aggregated results from the four pipelines
```{r Initialize RA results}
indivSig_TB_microarray_path <- here("results/microarray/04_rank_aggregation/RAresult_indivSig_TB_microarray.tsv")
aggSig_TB_microarray_path <- here("results/microarray/04_rank_aggregation/RAresult_neg_left_0.9_aggrSig_TB_microarray.tsv")
indivSig_TB_RNAseq_path <- here("results/RNAseq/04_rank_aggregation/RAresult_indivSig_TB_RNAseq.tsv")
aggSig_TB_RNAseq_path <- here("results/RNAseq/04_rank_aggregation/RAresult_neg_left_0.9_aggrSig_TB_RNAseq.tsv")
```

##### Prepare the pipeline results
```{r Read RA results}
pipeline_results <- read_pipeline_results(
    indivSig_TB_microarray_path,
    aggSig_TB_microarray_path,
    indivSig_TB_RNAseq_path,
    aggSig_TB_RNAseq_path
)

indivSig_TB_microarray <- pipeline_results$indivSig_TB_microarray
aggSig_TB_microarray <- pipeline_results$aggSig_TB_microarray
indivSig_TB_RNAseq <- pipeline_results$indivSig_TB_RNAseq
aggSig_TB_RNAseq <- pipeline_results$aggSig_TB_RNAseq

head(indivSig_TB_microarray, 5)
head(aggSig_TB_microarray, 5)
head(indivSig_TB_RNAseq, 5)
head(aggSig_TB_RNAseq, 5)
```

##### Get mean rank score of individual and aggregated drug predictions
```{r Process pipeline results}
pipeline_processed <- process_pipeline(
    indivSig_TB_microarray,
    indivSig_TB_RNAseq,
    aggSig_TB_microarray,
    aggSig_TB_RNAseq
)

indiv_results <- pipeline_processed$indiv_results
agg_results <- pipeline_processed$agg_results

head(indiv_results, 5)
head(agg_results, 5)
```

##### Get High Confidence Drug Predictions
```{r Get combined and high confidence predictions}
combined_predictions <- get_high_confidence_predictions(indiv_results, agg_results)[[1]]
high_confidence_predictions <- get_high_confidence_predictions(indiv_results, agg_results)[[2]]
```

```{r Save and display predictions}
output_path1 <- here("results/combined_drug_predictions.tsv")
save_and_display_predictions(combined_predictions, output_path1)

output_path2 <- here("results/final_high_confidence_drug_predictions.tsv")
save_and_display_predictions(high_confidence_predictions, output_path2)
```
